Storm surge modelling in the Strait of Georgia

### 2014-10-18 RoughWork/manuscript.tex
Mostly complete draft

\documentclass[pdftex,10pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}

@@

%storm surge modelling
%murty, danard, crean, non BC models (xu, flather)
Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructered grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion surface waves \citep{xu2010storm} have also been identified as important issues.  Most of these models neglect the effects of stratification, either by treating the ocean as a constant temperature and salinity or by using a two-dimensional depth-averaged set of equations, an assumption that is assumed to be valid.

@@

%outline
This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baraoclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep.  This paper is organized as follows: an overview of the model configuration and domain including a description of the ocean model we have employed is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents (K$_1$ ,O$_1$, P$_1$, Q$_1$, M$_2$, K$_2$, N$_2$, S$_2$), temperature and salinity climatologies, and the sea surface height anomaly. Tidal heights and currents at grid points along the Juan de Fuca boundary were extracted from Webtide, an online web prediction model for the northeast Pacific Ocean, which is based on work by \citet{foreman2000webtide}. The Johnstone Strait boundary was forced with current and elevation tidal harmonics measured and calculated by \citet{thomson1980johnstone} for the major M$_2$ and K$_1$ constituents. Additionally, O$_1$ and S$_2$ elevation harmonics from their measurements were employed. The remaining constituents were extrapolated from Webtide. At the western boundary, the tidal elevation and currents forcing parameters were tuned for a good match wiht observations at locations within the Strait of Georgia, including Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay. At the nothern boundary, the tides were tuned for good matches at Yorke Island and Kelsey Bay. Further details on the tuning are provided in the appendix \ref{sec:appendix}.

@@

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. Water level observations are taken from Fisheries and Oceans Canada (citation) and are used to calculate the osberved residual, defined as the observed water level minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this case, from the year prior surge event of interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

Several significant surge or wind events have been considered in detail, as well as an array of hindcasts to asses model ability in a statistical measure. These cases have been used to asses the model's skill in reproducing water level height and surge amplitude and timing during storm events. In addition, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed. We have also examined the spatial variability of the surge as it propagates through the domain and have considered a case with a strong wind event but no siginificant observed surge.

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is caclulated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill.

@@

The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations and corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. The timing of maximum water level matches well between the observations and the model at all four locations, the most significant contribution to high water level being from the tides which are generally well-represented by the model. There is a good match in the water level maximum at Victoria and Patricia Bay, with a relative error less than 2\%. There is only a 4.2 cm difference between observed and modelled water level elevation at Victoria and a 6.9 cm difference at Patricia Bay, the model slightly underpredicting the maximum water level in each case. The error at Point Atkinson and Campbell River is slightly more significant, reaching 2.5\% and 2.7\% respectively, close to 14 cm overpredicted in each case. It is suprising that Victoria and Patricia Bay agree most closely with observations because the M$_2$ tidal amplitudes in the model are known to be too low in this area. In conjunction with the overpredicted water levels at Point Atkinson and Campbell River, it is likely that one of the forcing factors is overestimated in this example.

A comparison between observed and modelled residuals is also displayed in the panels on the right. In general, there is a very good agreement between the observed and modelled residules. The most noticeable difference is a 4-6 hour delay in the model's timing of the maximum residual except at Victoria where the maximum residual in sync with observations. The errors in maximum residual range from 6.0 cm to high at Victoria to 15.9 cm too low at Campbell River. Point Atkinson and Patricia Bay both experience a very good match with observed residuals with errors less than 2 cm. Note that Point Atkinson and Campbell River observed the largest observed surges at 82 cm and 95 cm respectively. The surges at Victoria and Patricia Bay were observed to be 66 cm and 74 cm.

@@

The maximum water level and maximum residuals are important factors in storm surge hindcasts but a better indicator of overall model performance is provided by a group of statistical measures in Table \ref{tab:feb2006stat}. In general, $\gamma^2$ and $WS$ suggest good model agreement for the predicted water levels, most likely because the tidal signal dominates. The skill scores degrade slightly for the residuals but are still reasonable and indicate very good agreement. Note that the mean absolute error ($\bar{e}$) is lower for the residuals, but this measure does not account for the larger variablity in water levels due to the tides.

\begin{table}[h]
\centering 
\begin{tabular}{|l |c c c c c | c c c c c|} 
\hline 
& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
\hline 
Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
\hline 
Point Atkinson & 10.5           & 3.46                 & 3.54                 &   0.012    & 0.994  & 6.2            &  0.273                 & 0.309                  & 0.065     & 0.979     \\
Victoria       &  8.6           & 2.18                 & 2.19                 &   0.032    & 0.992  & 6.1            &  0.235                 & 0.278                  & 0.085     & 0.970     \\
Patricia Bay   &  9.1           & 2.60                 & 2.64                 &   0.024    & 0.992  & 6.9            &  0.223                 & 0.278                  & 0.070     & 0.969     \\
Campbell River & 13.1           & 3.20                 & 3.21                 &   0.041    & 0.990  & 5.8            &  0.303                 & 0.300                  & 0.079     & 0.961     \\
\hline 
\end{tabular}
\caption{Evaluation of model performance for the Feb 2006 simulation. Mean absolute error ($\bar{e}$), observed mean (subscript o), modelled mean (subscript m), $\gamma^2$, and the Willmott skill score ($WS$) are provided for water level predictions ($\eta$) and resdiuals ($\delta$) during the period displayed in Figure \ref{fig:feb2006}.}
\label{tab:feb2006stat} 
\end{table} 
%anything else to include? statistics calculated in Feb 2006 -paper prodcution.ipynb and saved in statisitcs_feb2006.csv 
%I also have standard deviations, correlations, ...

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. Interestingly, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are negelcted. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimtres if the inverse barometer effect due to atmospheric pressure is not included.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006.pdf}
\caption{Comparison of observed and modelled water level and residuls during the Dec 15, 2006 storm surge simulation. }
\label{fig:dec2006}
\end{figure}%table instead of figure? or maybe just the surge and not water level?

@@

The evolution and spatial distribution of the surge is highlighted in Figure \ref{fig:spatial} where the modelled residual is plotted every hour beginning on Dec 15 at 05:30 UTC through to Dec 15 at 14:30 UTC. The surge enters the system from the Strait of Juna de Fuca, peaking between Dec 15 at 9:30 and Dec 15 at 11:30 and then begins to subside throughout the entire domain. Notably, the Strait of Georgia experiences the largest surge values, upwards of 60 cm, whereas the Juan de Fuca region experiences a maximum surge around 45 cm. Additionally, the coastal areas on the mainland side of the Strait of Georgia experience slightly higher surges (\~ 10 cm) than the Vancouver Island side, likely due to the southerly wind in the atmospheric model pushing water against the mainland coast. At about 11:30 UTC, the winds switch to northwesterly and then westerly, further amplifying the surge along the mainland coast.

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006_spatial.pdf}
\caption{Propagation of the storm surge on Dec 15, 2006. Modelled residual every hour starting on Dec 15 at 05:30 UTC. Time advances from top left to bottom right. Wind vectors from the atmopsheric forcing are overlaid in black.}
\label{fig:spatial}
\end{figure}

An evaluation of the spatial distribution of the surge in a simulation without atmospheric forcing is provided in Figure \ref{fig:spatial_sshonly}. In this case, the surge amplitude across the Strait of Georgia, is uniform with no pronouced difference between the mainland side and Vancouver Island side. While the surge entering the domain from the Pacific Ocean is the leading order contribution, the winds may induce localized changes to the surge amplitude. Interestingly, the winds over the Strait of Juan de Fuca cause a decrease in surge amplitude in this region as the winds are acting to push the water towards the mainland.  Even without atmospheric forcing, there is still a siginificant difference in surge amplitude between the Strait of Georgria and the Strait of Juan de Fuca, perhaps because the Strait of Georgia is also affected by the surge entering the domain from Johnstone Strait at the northern boundary.

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006_spatial_sshonly.pdf}
\caption{As in Figure \ref{fig:spatial} but for a simulation that does not include atmopsheric forcing. }
\label{fig:spatial_sshonly}
\end{figure}

\subsection{Strong Wind Event: Nov 19, 2009}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. As recorded in the Environment Canada Historical Climate Data archives, the maximum hourly wind speed at Point Atkinson was 18.6 m/s and the largest gust observed was 25.8 m/s or 93 km/hr. In addition, large wind speeds over 15 m/s persisted for over 9 hours. Yet, the surge associated with this event is rather small when compared to the previous two cases as shown in Figure \ref{fig:nov2009} where we have plotted the wind speed at Point Atkinson, sea surface height anomaly forcing at the open boundaries, and observed and modelled residuals at Point Atkinson. Two significant surge events are noted, the first occurring on Nov 19 and a second, larger surge on Nov 20. The largest surge did not coincide with the strong wind event which peaked at Nov 19 01:00 UTC, reflected in both the model and observations. At Point Atkinsons, the highest observed surge reached 62.5 cm on Nov 20 at 10:30 UTC, while the model recorded a maximum surge of 50.5 cm five hours later. On Nov 19, the observed surge reached upwards of 50 cm, which is low when compared to the other cases considered. Neither of the surges occured during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy. 
%citation for Environment Canada data?

@@

\section{Discussion}\label{sec:diss}
In this article, we have configured a three-dimensional numerical ocean model of the Salish Sea for accurate hindcasts of storm surges in the Strait of Georgia. The model is forced with eight tidal constituents at the western and northern open boundaries, as well as sea surface height anomaly from de-tided observations at Tofino and Port Hardy. Additionaly, wind stress and an inverse barometer effect were forced at the ocean surface using an atmopsheric reforecasting product called CGRF \citep{smith2014new}. Storm surge simulations began with zero initial conditions for the currents and sea surface height and were commenced several days prior to the event of interest to spin up the velocity fields.

Hindcasts of significant events in February 2006, December 2006 and Novemeber 2009 were simulated. In general, the model performs well at reproducing the observed water level heights and surge amplitudes, with Willmott skill scores greater than 0.88 in all cases. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude (>50 \%), in agreement with previous studies in this region \citep{murty1995storm}. We have also found that the wind and atmospheric conditions can set up local differences in the surge amplitude, resulting in local variations in the spatial distribution of the surge. Generally, surge amplitudes in the Strait of Georgia are higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions. Results from an additional simulation with no tidal forcing but only surge propagation and atmopsheric forcing found that nonlinear effects due to tide-surge interaction were most important after the San Juan and Gulf Islands.

Several factors considered in other storm surge modelling studies have been neglected. Primarly, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accurcay are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Gerogia, such as Boundary Bay and Crescent Beach are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, sea surface height anomaly from Neah Bay will be used in future real time water leve predictions during the storm surge season.

The Strait of Georgia is a semi-enclosed body of water

@@

\appendix
\section{Appendix}\label{sec:appendix}
This section details the procedure used to tune the tidal forcing for a good match with observations in the Strait of Georgia. Originallly, tidal elevation and currents were extracted from a tide prediction tool called Webtide \citep{foreman2000webtide} at the western Juan de Fuca boundary. At the northern boundary,  M$_2$ and K$_1$ tidal harmonics for both elevation and currents were taken from \citet{thomson1980johnstone}, as well as elevation harmonics for O$_1$ and S$_2$. The remaining constituents were extrapolated from Webtide based on the ratio between M$_2$/S$_2$ amplitudes and phase differences for the semi-diurnal constituents and K$_1$/O$_1$ for the diurnal constituents. However, these forcing conditions did not result in a good match between observed and modelled tidal elevation harmonics in the Strait of Georgia, the main region of interest (although the matches in the Strait of Juan de Fuca were adequate). As such, tidal forcing was tuned, by first adjusting the western boundary M$_2$ anc K$_1$ phase and K$_1$ amplitude for a good match in the Strait of Georgia (specifically at Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay). This tuning was performed on simulations that used only M$_2$ and K$_1$ forcing. Next, the M$_2$ and K$_1$ fluxes at the northern boundary were adjusted because the cross-sectional area of transect used by \citet{thomson1980johnstone} to measure tidal currents in Johnstone Strait is smaller than the cross-sectional area of our northern boundary. Finally, the remaining constituents were added, starting with O$_1$, then S$_2$, and then in pairs P$_1$/N$_2$ and Q$_1$/K$_2$. Adjustments were made if the addition of new constituents hindered performance of the main M$_2$ and K$_1$ components.

>>>>

\documentclass[pdftex,10pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{setspace} \doublespacing

@@

%storm surge modelling
%murty, danard, crean, non BC models (xu, flather)
Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructured grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion surface waves \citep{xu2010storm} have also been identified as important issues.  Most of these models neglect the effects of stratification, either by treating the ocean as a constant temperature and salinity or by using a two-dimensional depth-averaged set of equations, an assumption that is assumed to be valid.

@@

%outline
This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baroclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep.  This paper is organized as follows: an overview of the model configuration and domain including a description of the ocean model we have employed is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents (K$_1$ ,O$_1$, P$_1$, Q$_1$, M$_2$, K$_2$, N$_2$, S$_2$), temperature and salinity climatologies, and the sea surface height anomaly. Tidal heights and currents at grid points along the Juan de Fuca boundary were extracted from Webtide, an online web prediction model for the northeast Pacific Ocean, which is based on work by \citet{foreman2000webtide}. The Johnstone Strait boundary was forced with current and elevation tidal harmonics measured and calculated by \citet{thomson1980johnstone} for the major M$_2$ and K$_1$ constituents. Additionally, O$_1$ and S$_2$ elevation harmonics from their measurements were employed. The remaining constituents were extrapolated from Webtide. At the western boundary, the tidal elevation and currents forcing parameters were tuned for a good match Whit observations at locations within the Strait of Georgia, including Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay. At the northern boundary, the tides were tuned for good matches at Yorke Island and Kelsey Bay. Further details on the tuning are provided in the appendix \ref{sec:appendix}.

@@

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. Water level observations are taken from Fisheries and Oceans Canada (citation) and are used to calculate the observed residual, defined as the observed water level minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this case, from the year prior surge event of interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

Several significant surge or wind events have been considered in detail, as well as an array of hindcasts to asses model ability in a statistical measure. These cases have been used to asses the model's skill in reproducing water level height and surge amplitude and timing during storm events. In addition, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed. We have also examined the spatial variability of the surge as it propagates through the domain and have considered a case with a strong wind event but no significant observed surge.

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill.

@@

The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations and corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. The timing of maximum water level matches well between the observations and the model at all four locations, the most significant contribution to high water level being from the tides which are generally well-represented by the model. There is a good match in the water level maximum at Victoria and Patricia Bay, with a relative error less than 2\%. There is only a 4.2 cm difference between observed and modelled water level elevation at Victoria and a 6.9 cm difference at Patricia Bay, the model slightly under predicting the maximum water level in each case. The error at Point Atkinson and Campbell River is slightly more significant, reaching 2.5\% and 2.7\% respectively, close to 14 cm over predicted in each case. It is sup rising that Victoria and Patricia Bay agree most closely with observations because the M$_2$ tidal amplitudes in the model are known to be too low in this area. In conjunction with the over predicted water levels at Point Atkinson and Campbell River, it is likely that one of the forcing factors is overestimated in this example.

A comparison between observed and modelled residuals is also displayed in the panels on the right. In general, there is a very good agreement between the observed and modelled residuals. The most noticeable difference is a 4-6 hour delay in the model's timing of the maximum residual except at Victoria where the maximum residual in sync with observations. The errors in maximum residual range from 6.0 cm to high at Victoria to 15.9 cm too low at Campbell River. Point Atkinson and Patricia Bay both experience a very good match with observed residuals with errors less than 2 cm. Note that Point Atkinson and Campbell River observed the largest observed surges at 82 cm and 95 cm respectively. The surges at Victoria and Patricia Bay were observed to be 66 cm and 74 cm.

@@

The maximum water level and maximum residuals are important factors in storm surge hindcasts but a better indicator of overall model performance is provided by a group of statistical measures in Table \ref{tab:feb2006stat}. In general, $\gamma^2$ and $WS$ suggest good model agreement for the predicted water levels, most likely because the tidal signal dominates. The skill scores degrade slightly for the residuals but are still reasonable and indicate very good agreement. Note that the mean absolute error ($\bar{e}$) is lower for the residuals, but this measure does not account for the larger variability in water levels due to the tides.

\begin{table}[h]
\centering 
\begin{tabular}{|l |c c c c c | c c c c c|} 
\hline 
& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
\hline 
Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
\hline 
Point Atkinson & 10.5           & 3.46                 & 3.54                 &   0.012    & 0.994  & 6.2            &  0.273                 & 0.309                  & 0.065     & 0.979     \\
Victoria       &  8.6           & 2.18                 & 2.19                 &   0.032    & 0.992  & 6.1            &  0.235                 & 0.278                  & 0.085     & 0.970     \\
Patricia Bay   &  9.1           & 2.60                 & 2.64                 &   0.024    & 0.992  & 6.9            &  0.223                 & 0.278                  & 0.070     & 0.969     \\
Campbell River & 13.1           & 3.20                 & 3.21                 &   0.041    & 0.990  & 5.8            &  0.303                 & 0.300                  & 0.079     & 0.961     \\
\hline 
\end{tabular}
\caption{Evaluation of model performance for the Feb 2006 simulation. Mean absolute error ($\bar{e}$), observed mean (subscript o), modelled mean (subscript m), $\gamma^2$, and the Willmott skill score ($WS$) are provided for water level predictions ($\eta$) and residuals ($\delta$) during the period displayed in Figure \ref{fig:feb2006}.}
\label{tab:feb2006stat} 
\end{table} 
%anything else to include? statistics calculated in Feb 2006 -paper production.ipynb and saved in statistics_feb2006.CV 
%I also have standard deviations, correlations, ...

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. Interestingly, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimetres if the inverse barometer effect due to atmospheric pressure is not included.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006.pdf}
\caption{Comparison of observed and modelled water level and residuals during the Dec 15, 2006 storm surge simulation. }
\label{fig:dec2006}
\end{figure}%table instead of figure? or maybe just the surge and not water level?

@@

The evolution and spatial distribution of the surge is highlighted in Figure \ref{fig:spatial} where the modelled residual is plotted every hour beginning on Dec 15 at 05:30 UTC through to Dec 15 at 14:30 UTC. The surge enters the system from the Strait of Juan de Fuca, peaking between Dec 15 at 9:30 and Dec 15 at 11:30 and then begins to subside throughout the entire domain. Notably, the Strait of Georgia experiences the largest surge values, upwards of 60 cm, whereas the Juan de Fuca region experiences a maximum surge around 45 cm. Additionally, the coastal areas on the mainland side of the Strait of Georgia experience slightly higher surges (\~ 10 cm) than the Vancouver Island side, likely due to the southerly wind in the atmospheric model pushing water against the mainland coast. At about 11:30 UTC, the winds switch to northwesterly and then westerly, further amplifying the surge along the mainland coast.

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006_spatial.pdf}
\caption{Propagation of the storm surge on Dec 15, 2006. Modelled residual every hour starting on Dec 15 at 05:30 UTC. Time advances from top left to bottom right. Wind vectors from the atmospheric forcing are overlaid in black.}
\label{fig:spatial}
\end{figure}

An evaluation of the spatial distribution of the surge in a simulation without atmospheric forcing is provided in Figure \ref{fig:spatial_sshonly}. In this case, the surge amplitude across the Strait of Georgia, is uniform with no pronounced difference between the mainland side and Vancouver Island side. While the surge entering the domain from the Pacific Ocean is the leading order contribution, the winds may induce localized changes to the surge amplitude. Interestingly, the winds over the Strait of Juan de Fuca cause a decrease in surge amplitude in this region as the winds are acting to push the water towards the mainland.  Even without atmospheric forcing, there is still a significant difference in surge amplitude between the Strait of Georgia and the Strait of Juan de Fuca, perhaps because the Strait of Georgia is also affected by the surge entering the domain from Johnstone Strait at the northern boundary.

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006_spatial_sshonly.pdf}
\caption{As in Figure \ref{fig:spatial} but for a simulation that does not include atmospheric forcing. }
\label{fig:spatial_sshonly}
\end{figure}

\subsection{Strong Wind Event: Nov 19, 2009}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. As recorded in the Environment Canada Historical Climate Data archives, the maximum hourly wind speed at Point Atkinson was 18.6 m/s and the largest gust observed was 25.8 m/s or 93 km/hr. In addition, large wind speeds over 15 m/s persisted for over 9 hours. Yet, the surge associated with this event is rather small when compared to the previous two cases as shown in Figure \ref{fig:nov2009} where we have plotted the wind speed at Point Atkinson, sea surface height anomaly forcing at the open boundaries, and observed and modelled residuals at Point Atkinson. Two significant surge events are noted, the first occurring on Nov 19 and a second, larger surge on Nov 20. The largest surge did not coincide with the strong wind event which peaked at Nov 19 01:00 UTC, reflected in both the model and observations. At Point Atkinson, the highest observed surge reached 62.5 cm on Nov 20 at 10:30 UTC, while the model recorded a maximum surge of 50.5 cm five hours later. On Nov 19, the observed surge reached upwards of 50 cm, which is low when compared to the other cases considered. Neither of the surges occurred during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy. 
%citation for Environment Canada data?

@@

\section{Discussion}\label{sec:diss}
In this article, we have configured a three-dimensional numerical ocean model of the Salish Sea for accurate hindcasts of storm surges in the Strait of Georgia. The model is forced with eight tidal constituents at the western and northern open boundaries, as well as sea surface height anomaly from de-tided observations at Tofino and Port Hardy. Additionally, wind stress and an inverse barometer effect were forced at the ocean surface using an atmospheric reforecasting product called CGRF \citep{smith2014new}. Storm surge simulations began with zero initial conditions for the currents and sea surface height and were commenced several days prior to the event of interest to spin up the velocity fields.

Hindcasts of significant events in February 2006, December 2006 and November 2009 were simulated. In general, the model performs well at reproducing the observed water level heights and surge amplitudes, with Willmott skill scores greater than 0.88 in all cases. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude (>50 \%), in agreement with previous studies in this region \citep{murty1995storm}. We have also found that the wind and atmospheric conditions can set up local differences in the surge amplitude, resulting in local variations in the spatial distribution of the surge. Generally, surge amplitudes in the Strait of Georgia are higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to tide-surge interaction were most important after the San Juan and Gulf Islands.

Several factors considered in other storm surge modelling studies have been neglected. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, sea surface height anomaly from Neah Bay will be used in future real time water level predictions during the storm surge season.

Since the Strait of Georgia is a large, semi-enclosed body of water storm surges can behave and evolve differently than in other coastal regions. To be continued...

@@

\appendix
\section{Appendix}\label{sec:appendix}
This section details the procedure used to tune the tidal forcing for a good match with observations in the Strait of Georgia. Originally, tidal elevation and currents were extracted from a tide prediction tool called Webtide \citep{foreman2000webtide} at the western Juan de Fuca boundary. At the northern boundary,  M$_2$ and K$_1$ tidal harmonics for both elevation and currents were taken from \citet{thomson1980johnstone}, as well as elevation harmonics for O$_1$ and S$_2$. The remaining constituents were extrapolated from Webtide based on the ratio between M$_2$/S$_2$ amplitudes and phase differences for the semi-diurnal constituents and K$_1$/O$_1$ for the diurnal constituents. However, these forcing conditions did not result in a good match between observed and modelled tidal elevation harmonics in the Strait of Georgia, the main region of interest (although the matches in the Strait of Juan de Fuca were adequate). As such, tidal forcing was tuned, by first adjusting the western boundary M$_2$ and K$_1$ phase and K$_1$ amplitude for a good match in the Strait of Georgia (specifically at Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay). This tuning was performed on simulations that used only M$_2$ and K$_1$ forcing. Next, the M$_2$ and K$_1$ fluxes at the northern boundary were adjusted because the cross-sectional area of transect used by \citet{thomson1980johnstone} to measure tidal currents in Johnstone Strait is smaller than the cross-sectional area of our northern boundary. Finally, the remaining constituents were added, starting with O$_1$, then S$_2$, and then in pairs P$_1$/N$_2$ and Q$_1$/K$_2$. Adjustments were made if the addition of new constituents hindered performance of the main M$_2$ and K$_1$ components.
